---
title: "Growth Modeling with fit_bayesian_growth()"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Growth Modeling with fit_bayesian_growth()}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 eval = FALSE,
 message = FALSE,
 warning = FALSE
)
```

## Overview

The `fit_bayesian_growth()` function fits von Bertalanffy, Gompertz, or Logistic growth models using Bayesian methods. A key innovation is the **maturity-based parameterization**, which derives the growth coefficient $k$ from observable maturity metrics rather than estimating it directly.

## Growth Model Equations

**von Bertalanffy** (`model = "vb"`):
$$L(t) = L_\infty - (L_\infty - L_0) e^{-kt}$$

**Gompertz** (`model = "gompertz"`):
$$L(t) = L_\infty \exp\left[-\ln\left(\frac{L_\infty}{L_0}\right) e^{-kt}\right]$$

**Logistic** (`model = "logistic"`):
$$L(t) = \frac{L_\infty}{1 + \left(\frac{L_\infty}{L_0} - 1\right) e^{-kt}}$$

## Two Parameterization Approaches

### 1. K-Based (Traditional)

Directly estimates the growth coefficient $k$:

```{r k-based}
library(vitalBayes)
library(data.table)

# Load simulated data
data(growth_data)

# Filter to non-embryos with age data
gdata <- growth_data[embryo == FALSE & !is.na(age)]

# Traditional k-based von Bertalanffy
growth_k <- fit_bayesian_growth(
 lt      = "fl",
 age     = "age",
 sex     = "sex",
 data    = gdata,
 model   = "vb",
 k_based = TRUE,
 CV_k    = 0.5,        # 50% CV on k prior (high uncertainty)
 parallel = TRUE
)

growth_k$summary(c("Linf", "L0", "k", "sigma"))
```

### 2. Maturity-Based (Recommended)

Derives $k$ from maturity parameters ($L_{mat}$, $t_{mat}$), ensuring biological consistency:

$$k = \frac{1}{t_{mat}} \ln\left(\frac{L_\infty - L_0}{L_\infty - L_{mat}}\right)$$

```{r mat-based}
# First, fit maturity models
mat_data <- growth_data[embryo == FALSE & !is.na(mat)]

L50_fit <- fit_bayesian_maturity(
 maturity = "mat", lt = "fl", sex = "sex",
 data = mat_data
)

t50_fit <- fit_bayesian_maturity(
 maturity = "mat", age = "age", sex = "sex",
 data = mat_data[!is.na(age)]
)

# Optional: fit birth model for L0 prior
birth_fit <- fit_bayesian_birth(
 embryo_lts = growth_data[embryo == TRUE, fl],
 free_swimming_lts = growth_data[embryo == FALSE, fl]
)

# Maturity-based growth model
growth_mat <- fit_bayesian_growth(
 lt        = "fl",
 age       = "age",
 sex       = "sex",
 data      = gdata,
 model     = "vb",
 k_based   = FALSE,          # Use maturity-based parameterization
 L50_fit   = L50_fit,        # Provides Lmat prior
 t50_fit   = t50_fit,        # Provides tmat prior
 birth_fit = birth_fit,      # Provides L0 prior
 parallel  = TRUE
)

# k is now a derived quantity
growth_mat$summary(c("Linf", "L0", "Lmat", "tmat", "k", "sigma"))
```

**Why maturity-based?**

1. **Reduced correlation**: $(L_\infty, k)$ are notoriously correlated; maturity parameters break this
2. **Observable anchoring**: Maturity milestones fall within the data range, unlike $L_\infty$
3. **Prior propagation**: Uncertainty from maturity models flows into growth estimates
4. **Biological coherence**: The growth curve *must* pass through $(t_{mat}, L_{mat})$

For the full derivation, see the [Maturity-Based Parameterization section](Understanding_vitalBayes.html#maturity-param) in the Statistical Methods guide.

## The $L_\infty$ Constraint

A critical issue: unconstrained $L_\infty$ often converges to values *below* the largest observed individuals—biologically impossible.

**vitalBayes enforces** $L_\infty > L_{max}$ (the maximum observed length). For the biological and statistical rationale, see [Why $L_\infty$ Must Exceed Maximum Observed Length](Understanding_vitalBayes.html#linf) in the Statistical Methods guide.

```{r linf}
# Lmax is auto-detected from data
growth_fit <- fit_bayesian_growth(
 lt   = "fl",
 age  = "age",
 data = gdata
)
# Message: "Lmax from data: 98.5 cm"

# Or specify manually (e.g., if you have length data without age)
growth_fit <- fit_bayesian_growth(
 lt   = "fl",
 age  = "age",
 data = gdata,
 Lmax = c(100, 95)  # Female, Male
)
```

## Two-Sex Models with Partial Pooling

When sample sizes are imbalanced between sexes (common in elasmobranch research), partial pooling borrows strength across sexes to reduce uncertainty for the sparse group:

```{r two-sex}
growth_2sex <- fit_bayesian_growth(
 lt          = "fl",
 age         = "age",
 sex         = "sex",
 data        = gdata,
 model       = "vb",
 k_based     = FALSE,
 L50_fit     = L50_fit,
 t50_fit     = t50_fit,
 use_pooling = TRUE,    # Partial pooling between sexes
 prior_tau   = 0.2,     # Half-normal scale for between-sex SD
 parallel    = TRUE
)

# Sex-specific estimates
growth_2sex$summary("Linf")
growth_2sex$summary("k")

# Sex differences
growth_2sex$summary(c("Linf_diff", "k_diff"))
```

For a comprehensive treatment of partial pooling mechanics, shrinkage effects, and when to use it, see `vignette("partial_pooling")`.

## Comparing Growth Models

```{r compare}
# Fit all three models
vb_fit <- fit_bayesian_growth(
 lt = "fl", age = "age", sex = "sex", data = gdata,
 model = "vb", k_based = FALSE, L50_fit = L50_fit, t50_fit = t50_fit
)

gomp_fit <- fit_bayesian_growth(
 lt = "fl", age = "age", sex = "sex", data = gdata,
 model = "gompertz", k_based = FALSE, L50_fit = L50_fit, t50_fit = t50_fit
)

logis_fit <- fit_bayesian_growth(
 lt = "fl", age = "age", sex = "sex", data = gdata,
 model = "logistic", k_based = FALSE, L50_fit = L50_fit, t50_fit = t50_fit
)

# Compare via LOO-CV
loo_vb <- compute_loo(vb_fit)
loo_gomp <- compute_loo(gomp_fit)
loo_logis <- compute_loo(logis_fit)

compare_loo(
 "von Bertalanffy" = loo_vb,
 "Gompertz" = loo_gomp,
 "Logistic" = loo_logis
)
```

## CV-Based Prior Specification

Priors are specified via coefficient of variation for intuitive scaling:

```{r cv-priors}
growth_fit <- fit_bayesian_growth(
 lt   = "fl",
 age  = "age",
 data = gdata,
 
 # Prior CVs (proportion of mean)
 CV_Linf = 0.20,    # 20% uncertainty on Linf
 CV_L0   = 0.30,    # 30% uncertainty on L0
 CV_k    = 0.50,    # 50% uncertainty on k (if k_based = TRUE)
 CV_Lmat = 0.20,    # 20% uncertainty on Lmat
 CV_tmat = 0.30,    # 30% uncertainty on tmat
 
 # Linf prior mean = 1.05 * Lmax by default
 Linf_multiplier = 1.05
)
```

## Visualization

```{r viz}
# Basic growth curve
plot_growth_curve(
 fit        = growth_2sex,
 data       = gdata,
 age_col    = "age",
 length_col = "fl",
 sex_col    = "sex"
)

# Multilingual support
plot_growth_curve(
 fit        = growth_2sex,
 data       = gdata,
 sex_labels = c("female" = "Hembra", "male" = "Macho"),
 x_lab      = "Edad (años)",
 y_lab      = "Longitud (cm)"
)

# Compare models visually
compare_growth_models(
 "von Bertalanffy" = vb_fit,
 "Gompertz" = gomp_fit,
 "Logistic" = logis_fit,
 data = gdata,
 age_col = "age",
 length_col = "fl"
)
```

## Posterior Predictive Checks

```{r ppc}
# Built-in PPC metrics
growth_2sex$summary(c("rmse", "mae", "prop_in_95ci"))

# Residual diagnostics
plot_residuals(
 fit        = growth_2sex,
 data       = gdata,
 age_col    = "age",
 length_col = "fl",
 type       = "all"
)
```

## Complete Workflow Example

```{r workflow}
# Load data
data(growth_data)

# ---- Stage 1: Birth ----
birth_fit <- fit_bayesian_birth(
 embryo_lts = growth_data[embryo == TRUE, fl],
 free_swimming_lts = growth_data[embryo == FALSE, fl]
)

# ---- Stage 2: Maturity ----
mat_data <- growth_data[embryo == FALSE & !is.na(mat)]

L50_fit <- fit_bayesian_maturity(
 maturity = "mat", lt = "fl", sex = "sex",
 data = mat_data, use_pooling = TRUE
)

t50_fit <- fit_bayesian_maturity(
 maturity = "mat", age = "age", sex = "sex",
 data = mat_data[!is.na(age)], use_pooling = TRUE
)

# ---- Stage 3: Growth ----
growth_fit <- fit_bayesian_growth(
 lt        = "fl",
 age       = "age",
 sex       = "sex",
 data      = growth_data[embryo == FALSE & !is.na(age)],
 model     = "vb",
 k_based   = FALSE,
 birth_fit = birth_fit,
 L50_fit   = L50_fit,
 t50_fit   = t50_fit,
 use_pooling = TRUE
)

# ---- Summary ----
create_parameter_table(
 birth = birth_fit,
 L50 = L50_fit,
 t50 = t50_fit,
 growth = growth_fit
)
```

## Troubleshooting

| Issue | Solution |
|-------|----------|
| Divergent transitions | Increase `adapt_delta` (0.95 → 0.99) |
| $k$ hitting boundaries | Check that $L_{mat} < L_\infty$ and $L_0 < L_{mat}$ |
| $L_\infty$ too low | Increase `Lmax` or `Linf_multiplier` |
| Poor fit at young ages | Consider different growth model (Gompertz often better for juveniles) |

## See Also

- `vignette("partial_pooling")` — When and how to use hierarchical structure for imbalanced sex ratios
- [Statistical Methods: Growth Models](Understanding_vitalBayes.html#growth) — Full mathematical derivation
- [Statistical Methods: Maturity-Based Parameterization](Understanding_vitalBayes.html#maturity-param) — The key innovation
- [Statistical Methods: CV-Based Priors](Understanding_vitalBayes.html#cv-priors) — How priors are specified
- `plot_growth_curve()`, `compare_growth_models()` — Visualization
- `compute_loo()`, `compare_loo()` — Model comparison

---

*This document is part of the vitalBayes R package. For bug reports, feature requests, or questions, please visit the GitHub repository.*
