---
title: "Estimating Birth Size with fit_bayesian_birth()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating Birth Size with fit_bayesian_birth()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 echo = TRUE,
 eval = FALSE,
 message = FALSE,
 warning = FALSE
)
```

## Overview
The `fit_bayesian_birth()` function estimates birth size ($b_{50}$) by comparing length distributions of embryos and free-swimming individuals using Bayesian probit regression.

**Key output**: The length at which 50% of individuals have transitioned from embryo to free-swimming status.

## The Model
The probability of being free-swimming follows a probit model:

$$P(\text{free-swimming}_i) = \Phi[\beta \cdot (L_i - b_{50})]$$

where:

- $\Phi(\cdot)$ is the standard normal CDF
- $b_{50}$ is the length at 50% transition probability
- $\beta > 0$ controls the sharpness of the transition

**Why probit?** The probit link assumes that underlying developmental readiness is normally distributed across individuals—a biologically intuitive interpretation. For a detailed derivation, see the [Probit Link section](Understanding_vitalBayes.html#probit) in the Statistical Methods guide.

## Basic Usage

```{r basic-usage}
library(vitalBayes)
library(data.table)

# Load example data (simulated von Bertalanffy growth data)
data(growth_data)

# Separate embryo and free-swimming lengths
embryo_lengths <- growth_data[embryo == TRUE, fl]
freeswim_lengths <- growth_data[embryo == FALSE, fl]

# Fit the birth model
birth_fit <- fit_bayesian_birth(
 embryo_lts        = embryo_lengths,
 free_swimming_lts = freeswim_lengths,
 chains            = 4,
 parallel          = TRUE
)
```

## Examining Results

```{r results}
# Core parameter estimates
birth_fit$summary(c("b50", "slope", "transition_width"))

# Posterior predictive checks
birth_fit$summary(c("mean_p_embryo", "mean_p_freeswim", "prop_correct_rep"))
```

**Key diagnostics to check:**

- `mean_p_embryo`: Should be low (~0.1-0.3) — embryos have low probability of being classified as free-swimming
- `mean_p_freeswim`: Should be high (~0.7-0.9) — free-swimmers have high probability
- `prop_correct_rep`: Overall classification accuracy (aim for >80%)

## Customizing Priors

The function automatically derives prior centers from the data, but you can override:
```{r priors}
birth_fit <- fit_bayesian_birth(
 embryo_lts        = embryo_lengths,
 free_swimming_lts = freeswim_lengths,
 mean_b50          = 35,      # Prior mean for b50 (cm)
 sd_b50            = 5,       # Prior SD
 mean_slope        = 0,       # Prior mean for log(slope)
 sd_slope          = 1,       # Prior SD for log(slope)
 adapt_delta       = 0.95     # Increase if divergences occur
)
```

## Visualization

```{r viz}
# Quick visualization
plot_birth_ogive(birth_fit, embryo_lengths, freeswim_lengths)

# Extract b50 for reporting
b50_summary <- birth_fit$summary("b50")
cat(sprintf("b50 = %.1f cm (95%% CI: %.1f – %.1f)\n",
           b50_summary$median, b50_summary$q5, b50_summary$q95))
```

## Using Output in Downstream Models

The $b_{50}$ estimate serves as the prior for birth length ($L_0$) in growth models:

```{r downstream}
# The birth fit can be passed directly to growth models
growth_fit <- fit_bayesian_growth(
 lt        = "fl",
 age       = "age", 
 sex       = "sex",
 data      = growth_data[embryo == FALSE],
 birth_fit = birth_fit  # Automatically extracts L0 prior
)
```

## Working with Limited Data

For datasets with few embryos, consider using the `limited_data` example dataset to explore prior sensitivity:

```{r limited}
# Load the small sample dataset (24F, 18M, 5 embryos)
data(limited_data)

# With sparse embryo data, priors become more influential
birth_fit_limited <- fit_bayesian_birth(
 embryo_lts        = limited_data[embryo == TRUE, fl],
 free_swimming_lts = limited_data[embryo == FALSE, fl],
 cv_b50            = 0.2  # Tighter prior when data are sparse
)
```

## Troubleshooting

| Issue | Solution |
|-------|----------|
| Divergent transitions | Increase `adapt_delta` (0.95 → 0.99) |
| Wide credible intervals | Check for data overlap; consider more informative priors |
| Poor classification | Verify embryo/free-swimming labels in data |

## See Also

- [Statistical Methods: Birth Size Estimation](Understanding_vitalBayes.html#birth) — Full mathematical derivation
- `plot_birth_ogive()` — Visualization function
- `fit_bayesian_maturity()` — Similar approach for maturity estimation
