---
title: "Age-Specific Natural Mortality with get_stochastic_mortality()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Age-Specific Natural Mortality with get_stochastic_mortality()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

## Overview

Natural mortality ($M$) is among the most difficult parameters to estimate in fisheries science, yet it profoundly influences population dynamics, sustainable harvest levels, and conservation assessments. For elasmobranchs, obtaining direct mortality estimates is particularly challenging, necessitating the need for reliable indirect estimation.

The `get_stochastic_mortality()` function generates Monte Carlo age-specific mortality schedules from growth model posteriors. By drawing parameter sets from the joint posterior distribution of a fitted growth model, the function propagates parameter uncertainty through to mortality estimates, producing credible intervals that honestly reflect our knowledge (or lack thereof) about the population. Three mortality models are available, each with distinct theoretical foundations, and all are compatible with any growth model fitted via `fit_bayesian_growth()`.

This vignette covers:

- Equations and biological rationale for each mortality model
- Scaling mortality schedules to empirical targets
- The Monte Carlo workflow and parameter handling
- Model comparison and sensitivity analysis
- Integration with downstream survival simulation

## Mortality Models

### Chen-Watanabe (1989)

The Chen-Watanabe model derives age-specific mortality from the relationship between body growth and mortality risk under von Bertalanffy dynamics:

$$M(t) = \frac{k_{vb} \cdot L_\infty}{L(t)}$$

where $k_{vb}$ is the von Bertalanffy growth coefficient serving as the asymptotic mortality rate ($M_\infty$), $L_\infty$ is asymptotic length, and $L(t)$ is predicted length at age $t$. Mortality is inversely proportional to body size: small, young individuals experience the highest mortality, which declines as the organism grows toward $L_\infty$.

This equation involves $k$ in two distinct roles. The $k_{vb}$ in the numerator is a *theoretical constant* derived from the CW formulation under von Bertalanffy assumptions --- it must always be the VB growth coefficient, regardless of which growth model was actually fitted. The $L(t)$ in the denominator is a *growth trajectory prediction* --- it should come from whatever model best describes the organism's actual growth dynamics. When a Gompertz or Logistic model is fitted, vitalBayes uses the native growth equation with native $k$ for $L(t)$, while computing a separate VB-equivalent $k$ for $M_\infty$. The VB-equivalent is derived from the shared biological milestones $(L_\infty, L_0, L_{mat}, t_{mat})$:

$$k_{vb}^{equiv} = \frac{1}{t_{mat}} \ln\!\left(\frac{L_\infty - L_0}{L_\infty - L_{mat}}\right)$$

For k-based fits without maturity milestones, a derivative-matching fallback is used (see *Derivative-Based Conversion* below).

The Chen-Watanabe model also supports a two-phase extension for late-life senescence, in which mortality transitions from the declining CW trajectory to an increasing senescence function (Gompertz or Logistic) at a fraction of the age at maturity. For the complete mathematical development of the CW reparameterization, the normalized growth coefficient $G(t)$, and the senescence extensions, see `vignette("chen_watanabe_reparameterization")`.

### Peterson-Wroblewski (1984)

The Peterson-Wroblewski model estimates mortality from an allometric relationship between body weight and mortality rate, derived from empirical data on pelagic fish populations:

$$M(W) = 1.92 \cdot W(t)^{-0.25}$$

where $W(t)$ is body weight in grams at age $t$, computed from predicted length via a user-supplied length-weight function $W = f(L)$. 

Unlike the CW model, the PW model requires a length-weight relationship function (provided via the `lw_fun` argument). Because the model operates on predicted body weight rather than growth model parameters directly, it is inherently growth-model-agnostic. Any growth model that predicts $L(t)$ can be combined with any $L$-$W$ conversion to produce mortality estimates.

### Lorenzen (1996, 2022)

The Lorenzen model relates natural mortality to body size through empirical calibration across a broad taxonomic range. vitalBayes supports two formulations.

**Weight-based** (Lorenzen, 1996):

$$M(W) = \alpha \cdot W(t)^{\beta}$$

where $\alpha$ and $\beta$ are allometric parameters estimated from a meta-analysis of marine and freshwater fishes ($\alpha \sim \mathcal{N}(3.69, \, 0.502)$, $\beta \sim \mathcal{N}(-0.305, \, 0.029)$). Like the PW model, $W(t)$ is calculated through a user-supplied length-weight function.

**Growth-based** (Lorenzen, 2022):

$$\ln M = 0.28 - 1.30 \, \ln\!\left(\frac{L(t)}{L_\infty}\right) + 1.08 \, \ln(k_{vb})$$

This formulation has the same dual-$k$ structure as Chen-Watanabe. The $\ln(k_{vb})$ term was calibrated against von Bertalanffy parameters (Lorenzen, 2022), so it must be the VB growth coefficient. The $L(t)/L_\infty$ ratio represents relative body size at age and should use the native growth trajectory for the most accurate prediction. When inputs come from a Gompertz or Logistic fit, vitalBayes passes the native $k$ for $L(t)$ prediction and a separate $k_{vb}^{equiv}$ for the regression term. The growth-based formulation has the advantage of not requiring a length-weight relationship, making it useful when $L$-$W$ data are unavailable.

### Derivative-Based $k_{vb}$ Conversion

When a k-based Gompertz or Logistic fit is used without maturity milestones, the VB-equivalent $k$ cannot be derived from $(L_{mat}, t_{mat})$. In this case, vitalBayes falls back to derivative matching at birth --- computing the VB growth coefficient that produces the same instantaneous growth rate at age 0:

$$k_{vb}^{equiv} = k_g \cdot \frac{L_0 \cdot \ln(L_\infty / L_0)}{L_\infty - L_0} \quad \text{(Gompertz)}$$

$$k_{vb}^{equiv} = k_l \cdot \frac{L_0}{L_\infty} \quad \text{(Logistic)}$$

These closed-form expressions follow from equating the VB growth rate at birth, $dL/dt|_{t=0} = k_{vb}(L_\infty - L_0)$, with the corresponding Gompertz or Logistic growth rate. The maturity milestone-based derivation is preferred when maturity data are available, since it anchors the conversion at a biologically meaningful interior point of the growth trajectory.

## Scaling Mortality Schedules

### Why Scale?

Theoretical mortality models (CW, PW, Lorenzen) produce age-specific *shapes* that are biologically informative, but their absolute levels often do not match empirical observations for elasmobranchs. This is expected as the models were derived from general relationships across broad taxonomic groups, not calibrated to the specific population under study. Scaling adjusts the overall mortality level while preserving the age-specific pattern, anchoring the curve to an independent empirical estimate.

### The Scaling Transformation

Scaling finds a proportional constant $c$ such that $M_{scaled}(t) = c \times M_{raw}(t)$. This multiplicative adjustment preserves the relative differences between ages --- if neonatal mortality is 5$\times$ adult mortality in the raw schedule, it remains 5$\times$ in the scaled schedule.

The constant $c$ is determined by matching the *cumulative hazard* of the scaled schedule to the target. The cumulative hazard $H = \int_0^{t_{max}} M(a) \, da$ is the quantity that directly governs survival: $S(t_{max}) = e^{-H}$. vitalBayes computes the raw cumulative hazard via trapezoidal numerical integration of the mortality schedule over the age grid:

$$H_{raw} = \int_0^{t_{max}} M_{raw}(a) \, da \approx \sum_{i=1}^{n-1} \frac{a_{i+1} - a_i}{2} \left[ M_{raw}(a_i) + M_{raw}(a_{i+1}) \right]$$

This approach is preferred over the simpler arithmetic-mean-based scaling ($c = \bar{M}_{target} / \bar{M}_{raw}$) because it directly constrains the biologically relevant quantity (cumulative survival probability) and is invariant to age grid spacing. 

### Target Specification

The scaling target can be specified in three ways, each producing a target cumulative hazard $H_{target}$ and scaling constant $c = H_{target} / H_{raw}$.

**Survival probability** (`M_target = NULL, p = ...`): Specify the proportion of a cohort expected to survive to maximum age $t_{max}$. The target cumulative hazard is:

$$H_{target} = -\ln(p)$$

A common choice is $p = 0.001$ (0.1% survival to $t_{max}$), which produces moderate scaling for long-lived species. This is the most directly interpretable target: the scaled schedule will produce exactly $S(t_{max}) = p$ under the cumulative hazard framework.

**Empirical relationship** (`M_target = function(tmax) ...`): Supply a function of $t_{max}$, such as the Then et al. (2015) estimator:

$$\bar{M}_{target} = 4.899 \cdot t_{max}^{-0.916}$$

This is interpreted as the average mortality rate over the lifespan, producing $H_{target} = \bar{M}_{target} \times t_{max}$.

**Fixed value** (`M_target = 0.15`): Directly specify a target mean mortality from literature estimates or independent analyses, also interpreted as the lifespan-averaged rate ($H_{target} = 0.15 \times t_{max}$).

```{r scaling-examples}
library(vitalBayes)

# Survival probability scaling (0.1% survive to tmax)
mort_p <- get_stochastic_mortality(
  method = "CW", growth_fit = growth_fit, sex = 1,
  scaled = TRUE, M_target = NULL, p = 0.001
)

# Then et al. (2015) empirical relationship
then_2015 <- function(tmax) 4.899 * tmax^(-0.916)
mort_then <- get_stochastic_mortality(
  method = "CW", growth_fit = growth_fit, sex = 1,
  scaled = TRUE, M_target = then_2015
)

# Fixed target from independent analysis
mort_fixed <- get_stochastic_mortality(
  method = "CW", growth_fit = growth_fit, sex = 1,
  scaled = TRUE, M_target = 0.15
)
```

## Basic Usage

### Standard Workflow with vitalBayes Fits

The typical workflow begins with growth (and optionally maturity) model fitting, then feeds posteriors into mortality estimation:

```{r basic-workflow}
library(vitalBayes)
library(data.table)

# Step 1: Fit growth model (any of the three)
# growth_fit <- fit_bayesian_growth(...)
# maturity_fit <- fit_bayesian_maturity(...)

# Step 2: Generate Chen-Watanabe mortality schedules
mort <- get_stochastic_mortality(
  method       = "CW",
  growth_fit   = growth_fit,
  maturity_fit = maturity_fit,
  sex          = 2,        # Males
  iter         = 2000,     # Number of posterior draws
  scaled       = TRUE,
  p            = 0.001,    # 0.1% survival to tmax
  two_phase    = TRUE,     # Include senescence
  late_model   = "gompertz",
  print_plot   = TRUE
)
```

The function automatically detects the growth model type from the fit object and extracts $(L_\infty, L_0, L_{mat}, t_{mat})$ from the joint posterior. When the growth model is Gompertz or Logistic, two things happen: the native $k$ and growth equation are used for $L(t)$ prediction, and a separate VB-equivalent $k$ is computed for mortality models that require it (CW and growth-based Lorenzen). Joint posterior sampling preserves parameter correlations.

### Output Structure

```{r output}
names(mort)
#> [1] "Schedules"  "Parameters" "Summary"    "Plot"

# Full age-specific schedules for every posterior draw
head(mort$Schedules)
#>    set_id  age     M_raw  M_scaled age_round
#>     <int> <num>     <num>    <num>     <num>
#> 1:      1   0.5  0.42318  0.31204       0.5
#> 2:      1   1.0  0.33721  0.24870       1.0

# Parameter draws used (includes both k values and growth model type)
head(mort$Parameters)
#>    set_id   Linf     L0   Lmat  tmat k_native k_vb_equiv growth_model  tmax
#>     <int>  <num>  <num>  <num> <num>    <num>      <num>       <char> <num>
#> 1:      1  98.42  24.31  68.92  11.2   0.1521     0.0945     gompertz  34.2

# Summary statistics across posterior draws
head(mort$Summary)
#>    age_round M_median M_mean M_lower M_upper
#>        <num>    <num>  <num>   <num>   <num>
#> 1:       0.5   0.3012 0.3124  0.2018  0.4532
```

### Manual Parameter Specification

For literature-based analyses where posterior draws are unavailable, parameters can be specified as mean-SD pairs. Two sampling modes are supported, mirroring the k-based and maturity-based parameterizations in `fit_bayesian_growth()`.

#### k-Based (Independent Sampling)

When only `k` is provided without `Lmat`, all parameters are sampled independently from their specified normal distributions. The `k` value should be the native growth coefficient for the model specified by `growth_model`. The function automatically derives $k_{vb}^{equiv}$ via derivative matching at birth and uses the native growth equation for $L(t)$ prediction.

```{r manual-k-vb}
# VB parameters (backward-compatible: k IS the VB k)
mort_vb <- get_stochastic_mortality(
  method = "CW",
  Linf   = c(100, 8),     # mean, sd
  L0     = c(25, 2),
  k      = c(0.08, 0.01), # VB k
  tmat   = c(12, 1.5),
  iter   = 2000,
  scaled = TRUE,
  p      = 0.001
)

# Gompertz parameters (derivative matching for k_vb)
mort_gomp <- get_stochastic_mortality(
  method       = "CW",
  Linf         = c(100, 8),
  L0           = c(25, 2),
  k            = c(0.15, 0.02),  # Gompertz k (NOT VB k)
  tmat         = c(12, 1.5),
  growth_model = "gompertz",     # Must match source of k
  iter         = 2000,
  scaled       = TRUE,
  p            = 0.001
)
```

Independent sampling destroys any correlations between parameters. This typically produces wider (less precise) uncertainty bands than joint posterior sampling from a fitted model, and can occasionally generate parameter combinations that are individually plausible but jointly inconsistent.

#### Maturity-Based (Bivariate Sampling)

When both `Lmat` and `tmat` are provided, they are sampled jointly from a bivariate normal distribution via Cholesky factorization. This preserves the positive biological correlation between length- and age-at-maturity: larger individuals tend to mature at older ages. The VB-equivalent $k$ is then *derived* from the sampled milestones $(L_\infty, L_0, L_{mat}, t_{mat})$ rather than taken from the user directly.

```{r manual-mat}
mort_mat <- get_stochastic_mortality(
  method        = "CW",
  Linf          = c(100, 8),
  L0            = c(25, 2),
  Lmat          = c(72, 4),       # length-at-maturity: mean, sd
  tmat          = c(12, 1.5),     # age-at-maturity: mean, sd
  k             = c(0.15, 0.02),  # Native Gompertz k (for L(t) trajectory)
  growth_model  = "gompertz",
  rho_Lmat_tmat = 0.5,            # Lmat-tmat correlation (default)
  iter          = 2000,
  scaled        = TRUE,
  p             = 0.001
)
```

The `rho_Lmat_tmat` argument controls the strength of the assumed correlation. The default of 0.5 reflects the broad biological expectation that larger individuals mature later. Alternative values can be obtained from paired individual-level data, from computing the empirical correlation between aligned posteriors of separate maturity model fits, from published species-specific estimates, or by running sensitivity analyses across a range of $\rho$ values.

When both `Lmat`/`tmat` and `k` are provided, the milestone-derived $k_{vb}^{equiv}$ is used for CW and growth-based Lorenzen (as the $M_\infty$ constant), while the user-supplied `k` is retained as the native growth coefficient for $L(t)$ prediction under the specified `growth_model`. This is the preferred configuration for non-VB growth models: the maturity milestones anchor the VB-equivalent $k$ at a biologically meaningful point, and the native $k$ produces the most accurate growth trajectory.

If `k` is omitted in this mode, the milestone-derived $k_{vb}^{equiv}$ is used for *both* roles, and $L(t)$ defaults to the VB equation regardless of `growth_model`. This is appropriate when only maturity data are available, but the function will issue a warning for non-VB `growth_model` settings.


## Comparing Mortality Models

### Across Mortality Methods

Comparing mortality estimates from different theoretical frameworks provides a measure of structural uncertainty:

```{r compare-methods}
# Chen-Watanabe
mort_cw <- get_stochastic_mortality(
  method = "CW", growth_fit = growth_fit, sex = 1,
  scaled = TRUE, p = 0.001, print_plot = FALSE
)

# Peterson-Wroblewski (requires length-weight function)
lw_fun <- function(L) 0.0001 * L^3.1
mort_pw <- get_stochastic_mortality(
  method = "PW", growth_fit = growth_fit, sex = 1,
  lw_fun = lw_fun, scaled = TRUE, p = 0.001,
  print_plot = FALSE
)

# Lorenzen growth-based
mort_lor <- get_stochastic_mortality(
  method = "L", growth_fit = growth_fit, sex = 1,
  weight_based = FALSE, scaled = TRUE, p = 0.001,
  print_plot = FALSE
)

# Combine for comparison plot
library(ggplot2)
combined <- rbind(
  mort_cw$Summary[, model := "Chen-Watanabe"],
  mort_pw$Summary[, model := "Peterson-Wroblewski"],
  mort_lor$Summary[, model := "Lorenzen"]
)

ggplot(combined, aes(x = age_round, color = model, fill = model)) +
  geom_ribbon(aes(ymin = M_lower, ymax = M_upper), alpha = 0.2, color = NA) +
  geom_line(aes(y = M_median), linewidth = 1) +
  scale_color_manual(values = vital_palette(3)) +
  scale_fill_manual(values = vital_palette(3)) +
  labs(x = "Age (years)", y = "Natural Mortality (M)",
       title = "Mortality Model Comparison") +
  theme_bw() +
  theme(legend.position = "top")
```

All three models predict declining mortality with age, but they differ in their curvature, early-age magnitude, and late-life behavior. Comparing across models gives a sense of how sensitive management conclusions are to the choice of theoretical framework.

### Across Growth Models

A complementary diagnostic tests whether mortality estimates are robust to the choice of growth model:

```{r compare-growth}
# Fit all three growth models
# vb_fit   <- fit_bayesian_growth(..., model = "v")
# gomp_fit <- fit_bayesian_growth(..., model = "g")
# logis_fit <- fit_bayesian_growth(..., model = "l")

# CW mortality from each growth model (auto-detects model type)
mort_from_vb <- get_stochastic_mortality(
  method = "CW", growth_fit = vb_fit, sex = 1, print_plot = FALSE
)
mort_from_gomp <- get_stochastic_mortality(
  method = "CW", growth_fit = gomp_fit, sex = 1, print_plot = FALSE
)
mort_from_logis <- get_stochastic_mortality(
  method = "CW", growth_fit = logis_fit, sex = 1, print_plot = FALSE
)

combined_growth <- rbind(
  mort_from_vb$Summary[, growth_model := "von Bertalanffy"],
  mort_from_gomp$Summary[, growth_model := "Gompertz"],
  mort_from_logis$Summary[, growth_model := "Logistic"]
)

ggplot(combined_growth, aes(x = age_round, color = growth_model, fill = growth_model)) +
  geom_ribbon(aes(ymin = M_lower, ymax = M_upper), alpha = 0.2, color = NA) +
  geom_line(aes(y = M_median), linewidth = 1) +
  labs(x = "Age (years)", y = "Natural Mortality (M)",
       title = "CW Mortality by Growth Model",
       subtitle = "Native L(t) trajectories with VB-equivalent M_inf") +
  theme_bw() +
  theme(legend.position = "top")
```

Differences between growth models in CW mortality now arise from two sources: the VB-equivalent $k$ (which may differ across models if the milestone-based inversions produce slightly different values), and --- more importantly --- the native $L(t)$ trajectories, which have different curvatures between anchor points. Models with faster early growth predict larger $L(t)$ at young ages and therefore lower juvenile mortality, while models that approach $L_\infty$ less tightly may produce slightly different late-age behavior. If the three curves are similar, mortality estimates are robust to growth model choice. Divergence indicates genuine model uncertainty that should be reported.

## Modular Mortality Functions

For applications outside the Monte Carlo framework vitalBayes exports the individual mortality model functions:

```{r modular}
ages <- seq(0.5, 30, by = 0.5)

# Chen-Watanabe with VB (k serves both roles)
M_cw_vb <- M_chen_watanabe_L0(
  age = ages, Linf = 100, L0 = 25, k = 0.08,
  two_phase = TRUE, tmat = 12, late_model = "gompertz"
)

# Chen-Watanabe with Gompertz (separated k roles)
M_cw_gomp <- M_chen_watanabe_L0(
  age = ages, Linf = 100, L0 = 25,
  k = 0.08,              # VB-equivalent k (for M_inf)
  k_native = 0.15,       # Gompertz k (for L(t))
  growth_model = "gompertz",
  two_phase = TRUE, tmat = 12
)

# Peterson-Wroblewski (native k only, no VB theory)
lw_func <- function(L) 0.0001 * L^3.1
M_pw <- M_peterson_wroblewski(
  age = ages, Linf = 100, L0 = 25, k = 0.15,
  lw_fun = lw_func, growth_model = "gompertz"
)

# Lorenzen growth-based (both k values)
M_lor <- M_lorenzen(
  age = ages, Linf = 100, L0 = 25,
  k = 0.15,          # Gompertz k (for L(t)/Linf ratio)
  k_vb = 0.08,       # VB-equivalent (for ln(k) regression term)
  weight_based = FALSE,
  growth_model = "gompertz"
)

# Lorenzen weight-based (native k only)
M_lor_wt <- M_lorenzen(
  age = ages, Linf = 100, L0 = 25, k = 0.15,
  lw_fun = lw_func, weight_based = TRUE,
  growth_model = "gompertz"
)

# Scale any of them
M_scaled <- scale_mortality(M_cw_gomp, M_target = NULL, tmax = 35, p = 0.001)
```

## Downstream Integration: Survival Simulation

Mortality schedules from `get_stochastic_mortality()` feed directly into `simulate_survivorship()` for cohort survival analysis:

```{r downstream}
surv <- simulate_survivorship(
  mc_object = mort,
  n         = 50000,    # Starting cohort size
  n_iter    = 2000,     # Simulation iterations
  mode      = "random"  # Sample parameter sets randomly
)

# Key survival metrics
surv$Aggregate$Age_of_Death
#>     mean    sd  lower  upper
#>    <num> <num>  <num>  <num>
#> 1:  8.42  2.31   4.89  13.21

surv$Aggregate$Survival_to_tmat
#>     mean      sd   lower   upper
#>    <num>   <num>   <num>   <num>
#> 1: 0.312  0.0891   0.168   0.512
```

See `vignette("survivorship_simulation")` for comprehensive coverage of survival analysis options, visualization, and interpretation.

## Troubleshooting

| Issue | Possible Cause | Solution |
|-------|----------------|----------|
| Negative mortality values | CW two-phase Taylor expansion breakdown | Function automatically caps at 0; consider single-phase or different `late_model` |
| Very wide uncertainty bands | Weak growth model posterior | Check growth model convergence and parameter correlations |
| Unrealistic tmax estimates | `Linf_factor` too extreme | Reduce from 0.99 to 0.95 for exploratory analysis |
| Missing tmat error | CW needs age-at-maturity | Provide `maturity_fit`, manual `tmat`, or both `Lmat` and `tmat` for bivariate sampling |
| `lw_fun` error | PW/Lorenzen weight-based needs length-weight | Provide function: `lw_fun = function(L) a * L^b` |
| CW and PW disagree strongly | Different theoretical assumptions | Report both; consider which assumptions best match your species |


## References

Chen, S., & Watanabe, S. (1989). Age dependence of natural mortality coefficient in fish population dynamics. *Nippon Suisan Gakkaishi*, 55(2), 205--208.

Lorenzen, K. (1996). The relationship between body weight and natural mortality in juvenile and adult fish. *Journal of Fish Biology*, 49(4), 627--642.

Lorenzen, K. (2022). Size- and age-dependent natural mortality in fish populations. *Fisheries Research*, 255, 106454.

Peterson, I., & Wroblewski, J. S. (1984). Mortality rate of fishes in the pelagic ecosystem. *Canadian Journal of Fisheries and Aquatic Sciences*, 41(7), 1117--1120.

Then, A. Y., Hoenig, J. M., Hall, N. G., & Hewitt, D. A. (2015). Evaluating the predictive performance of empirical estimators of natural mortality rate using information on over 200 fish species. *ICES Journal of Marine Science*, 72(1), 82--92.

## See Also

- `vignette("chen_watanabe_reparameterization")` --- Mathematical derivation of the CW reparameterization and $G(t)$ framework
- `vignette("survivorship_simulation")` --- Cohort survival analysis using mortality schedules
- `vignette("fit_bayesian_growth")` --- Growth model fitting that feeds the mortality workflow
- `vignette("fit_bayesian_maturity")` --- Maturity estimation for $t_{mat}$ and $L_{mat}$

---

*This document is part of the vitalBayes R package. For bug reports, feature requests, or questions, please visit the GitHub repository.*
