---
title: "Age-Specific Natural Mortality with get_stochastic_mortality()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Age-Specific Natural Mortality with get_stochastic_mortality()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

## Overview

Natural mortality ($M$) is among the most difficult parameters to estimate in fisheries science, yet it profoundly influences population dynamics, sustainable harvest levels, and conservation assessments. For elasmobranchs, obtaining direct mortality estimates is particularly challenging: long lifespans make mark-recapture studies impractical, populations are often too sparse for reliable catch-curve analysis, and the absence of otoliths necessitates alternative aging structures with their own uncertainties.

The `get_stochastic_mortality()` function addresses these challenges by generating Monte Carlo age-specific mortality schedules from growth model posteriors. By drawing parameter sets from the joint posterior distribution of a fitted growth model, the function propagates parameter uncertainty through to mortality estimates, producing credible intervals that honestly reflect our knowledge (or lack thereof) about the population. Three mortality models are available, each with distinct theoretical foundations, and all are compatible with any growth model fitted via `fit_bayesian_growth()`.

This vignette covers:

- Equations and biological rationale for each mortality model
- Scaling mortality schedules to empirical targets
- The Monte Carlo workflow and parameter handling
- Model comparison and sensitivity analysis
- Integration with downstream survival simulation

## Mortality Models

### Chen-Watanabe (1989)

The Chen-Watanabe model derives age-specific mortality from the relationship between body growth and mortality risk under von Bertalanffy dynamics. In the $L_0$-parameterized form used by vitalBayes, the model is:

$$M(t) = \frac{M_\infty \cdot L_\infty}{L(t)}$$

where $M_\infty \equiv k_{VB}$ is the asymptotic mortality rate (equal to the von Bertalanffy growth coefficient), $L_\infty$ is asymptotic length, and $L(t)$ is the predicted length at age $t$. Mortality is inversely proportional to body size: small, young individuals experience the highest mortality, which declines as the organism grows toward $L_\infty$.

When growth is fitted using a Gompertz or Logistic model rather than von Bertalanffy, the function automatically computes the VB-equivalent $k$ from the shared biological milestones $(L_\infty, L_0, L_{mat}, t_{mat})$:

$$k_{VB}^{equiv} = \frac{1}{t_{mat}} \ln\!\left(\frac{L_\infty - L_0}{L_\infty - L_{mat}}\right)$$

The Chen-Watanabe model also supports a two-phase extension for late-life senescence, in which mortality transitions from the declining CW trajectory to an increasing senescence function (Gompertz or Logistic) at a fraction of the age at maturity. For the complete mathematical development of the CW reparameterization, the normalized growth coefficient $G(t)$, and the senescence extensions, see `vignette("chen_watanabe_reparameterization")`.

### Peterson-Wroblewski (1984)

The Peterson-Wroblewski model estimates mortality from an allometric relationship between body weight and mortality rate, derived from empirical data on pelagic fish populations:

$$M(W) = 1.92 \cdot W(t)^{-0.25}$$

where $W(t)$ is body weight in grams at age $t$, computed from predicted length via a user-supplied length-weight function $W = f(L)$. The $-0.25$ exponent reflects metabolic scaling theory: metabolic rate scales approximately as $W^{0.75}$ (Kleiber's law), and mortality is assumed proportional to mass-specific metabolic rate, yielding a $W^{-0.25}$ dependence.

Unlike the CW model, the PW model requires a length-weight relationship function (provided via the `lw_fun` argument). Because the model operates on predicted body weight rather than growth model parameters directly, it is inherently growth-model-agnostic. Any growth model that predicts $L(t)$ can be combined with any $L$-$W$ conversion to produce mortality estimates. The length-at-age predictions use the native growth model's $k$ (not a VB-equivalent), since the PW model was not derived under VB-specific assumptions.

### Lorenzen (1996, 2022)

The Lorenzen model relates natural mortality to body size through empirical calibration across a broad taxonomic range. vitalBayes supports two formulations.

**Weight-based** (Lorenzen, 1996):

$$M(W) = \alpha \cdot W(t)^{\beta}$$

where $\alpha$ and $\beta$ are allometric parameters estimated from a meta-analysis of marine and freshwater fishes ($\alpha \sim \mathcal{N}(3.69, \, 0.502)$, $\beta \sim \mathcal{N}(-0.305, \, 0.029)$). Like the PW model, this requires a length-weight function and uses the native growth model's $k$ and $L(t)$ predictions.

**Growth-based** (Lorenzen, 2022):

$$\ln M = 0.28 - 1.30 \, \ln\!\left(\frac{L(t)}{L_\infty}\right) + 1.08 \, \ln(k)$$

This formulation was calibrated using von Bertalanffy parameters, so $k$ should be the VB-equivalent growth coefficient when inputs come from Gompertz or Logistic model fits. The growth-based formulation has the advantage of not requiring a length-weight relationship, making it useful when $L$-$W$ data are unavailable.

## Scaling Mortality Schedules

### Why Scale?

Theoretical mortality models (CW, PW, Lorenzen) produce age-specific *shapes* that are biologically informative, but their absolute levels often do not match empirical observations. This is expected: the models were derived from general relationships across broad taxonomic groups, not calibrated to the specific population under study. Scaling adjusts the overall mortality level while preserving the age-specific pattern, anchoring the curve to an independent empirical estimate.

### The Scaling Transformation

Given a raw mortality schedule $M_{raw}(t)$ and a target mean mortality $\bar{M}_{target}$, the scaled schedule is:

$$M_{scaled}(t) = M_{raw}(t) \times \frac{\bar{M}_{target}}{\bar{M}_{raw}}$$

where $\bar{M}_{raw} = \frac{1}{n} \sum_t M_{raw}(t)$ is the mean of the unscaled schedule. This multiplicative adjustment preserves the relative differences between ages --- if neonatal mortality is 5$\times$ adult mortality in the raw schedule, it remains 5$\times$ in the scaled schedule.

### Target Specification

The scaling target $\bar{M}_{target}$ can be specified in three ways.

**Survival probability** (`M_target = NULL, p = ...`): Specify the proportion of a cohort expected to survive to maximum age $t_{max}$. The target is derived as:

$$\bar{M}_{target} = -\frac{\ln(p)}{t_{max}}$$

A common choice is $p = 0.001$ (0.1% survival to $t_{max}$), which produces moderate scaling for long-lived species.

**Empirical relationship** (`M_target = function(tmax) ...`): Supply a function of $t_{max}$, such as the Then et al. (2015) estimator:

$$\bar{M}_{target} = 4.899 \cdot t_{max}^{-0.916}$$

This approach leverages the strong empirical correlation between maximum age and mean mortality across fish taxa.

**Fixed value** (`M_target = 0.15`): Directly specify a target mean mortality from literature estimates or independent analyses.

```{r scaling-examples}
library(vitalBayes)

# Survival probability scaling (0.1% survive to tmax)
mort_p <- get_stochastic_mortality(
  method = "CW", growth_fit = growth_fit, sex = 1,
  scaled = TRUE, M_target = NULL, p = 0.001
)

# Then et al. (2015) empirical relationship
then_2015 <- function(tmax) 4.899 * tmax^(-0.916)
mort_then <- get_stochastic_mortality(
  method = "CW", growth_fit = growth_fit, sex = 1,
  scaled = TRUE, M_target = then_2015
)

# Fixed target from independent analysis
mort_fixed <- get_stochastic_mortality(
  method = "CW", growth_fit = growth_fit, sex = 1,
  scaled = TRUE, M_target = 0.15
)
```

## Basic Usage

### Standard Workflow with vitalBayes Fits

The typical workflow begins with growth (and optionally maturity) model fitting, then feeds posteriors into mortality estimation:

```{r basic-workflow}
library(vitalBayes)
library(data.table)

# Step 1: Fit growth model (any of the three)
# growth_fit <- fit_bayesian_growth(...)
# maturity_fit <- fit_bayesian_maturity(...)

# Step 2: Generate Chen-Watanabe mortality schedules
mort <- get_stochastic_mortality(
  method       = "CW",
  growth_fit   = growth_fit,
  maturity_fit = maturity_fit,
  sex          = 2,        # Males
  iter         = 2000,     # Number of posterior draws
  scaled       = TRUE,
  p            = 0.001,    # 0.1% survival to tmax
  two_phase    = TRUE,     # Include senescence
  late_model   = "gompertz",
  print_plot   = TRUE
)
```

The function automatically extracts $(L_\infty, L_0, L_{mat}, t_{mat})$ from the growth posterior and computes the VB-equivalent $k$ when the growth model is Gompertz or Logistic. Joint posterior sampling preserves parameter correlations --- critically, the strong negative correlation between $L_\infty$ and $k$ that is typical of growth model fits.

### Output Structure

```{r output}
names(mort)
#> [1] "Schedules"  "Parameters" "Summary"    "Plot"

# Full age-specific schedules for every posterior draw
head(mort$Schedules)
#>    set_id  age     M_raw  M_scaled age_round
#>     <int> <num>     <num>    <num>     <num>
#> 1:      1   0.5  0.42318  0.31204       0.5
#> 2:      1   1.0  0.33721  0.24870       1.0

# Parameter draws used (includes k_vb_equiv)
head(mort$Parameters)
#>    set_id   Linf     L0   Lmat  tmat k_original k_vb_equiv k_for_mort  tmax
#>     <int>  <num>  <num>  <num> <num>      <num>      <num>      <num> <num>
#> 1:      1  98.42  24.31  68.92  11.2     0.0823     0.0945     0.0945  34.2

# Summary statistics across posterior draws
head(mort$Summary)
#>    age_round M_median M_mean M_lower M_upper
#>        <num>    <num>  <num>   <num>   <num>
#> 1:       0.5   0.3012 0.3124  0.2018  0.4532
```

### Manual Parameter Specification

For literature-based analyses where posterior draws are unavailable, parameters can be specified as mean-SD pairs:

```{r manual}
mort_manual <- get_stochastic_mortality(
  method = "CW",
  Linf   = c(100, 8),     # mean, sd
  L0     = c(25, 2),
  k      = c(0.08, 0.01),
  tmat   = c(12, 1.5),
  iter   = 2000,
  scaled = TRUE,
  p      = 0.001
)
```

Note that manual specification samples parameters independently, destroying any correlations. This typically produces wider (less precise) uncertainty bands than joint posterior sampling from a fitted model.

## Comparing Mortality Models

### Across Mortality Methods

Comparing mortality estimates from different theoretical frameworks provides a measure of structural uncertainty:

```{r compare-methods}
# Chen-Watanabe
mort_cw <- get_stochastic_mortality(
  method = "CW", growth_fit = growth_fit, sex = 1,
  scaled = TRUE, p = 0.001, print_plot = FALSE
)

# Peterson-Wroblewski (requires length-weight function)
lw_fun <- function(L) 0.0001 * L^3.1
mort_pw <- get_stochastic_mortality(
  method = "PW", growth_fit = growth_fit, sex = 1,
  lw_fun = lw_fun, scaled = TRUE, p = 0.001,
  print_plot = FALSE
)

# Lorenzen growth-based
mort_lor <- get_stochastic_mortality(
  method = "L", growth_fit = growth_fit, sex = 1,
  weight_based = FALSE, scaled = TRUE, p = 0.001,
  print_plot = FALSE
)

# Combine for comparison plot
library(ggplot2)
combined <- rbind(
  mort_cw$Summary[, model := "Chen-Watanabe"],
  mort_pw$Summary[, model := "Peterson-Wroblewski"],
  mort_lor$Summary[, model := "Lorenzen"]
)

ggplot(combined, aes(x = age_round, color = model, fill = model)) +
  geom_ribbon(aes(ymin = M_lower, ymax = M_upper), alpha = 0.2, color = NA) +
  geom_line(aes(y = M_median), linewidth = 1) +
  scale_color_manual(values = vital_palette(3)) +
  scale_fill_manual(values = vital_palette(3)) +
  labs(x = "Age (years)", y = "Natural Mortality (M)",
       title = "Mortality Model Comparison") +
  theme_bw() +
  theme(legend.position = "top")
```

All three models predict declining mortality with age, but they differ in their curvature, early-age magnitude, and late-life behavior. Comparing across models gives a sense of how sensitive management conclusions are to the choice of theoretical framework.

### Across Growth Models

A complementary diagnostic tests whether mortality estimates are robust to the choice of growth model:

```{r compare-growth}
# Fit all three growth models
# vb_fit   <- fit_bayesian_growth(..., model = "v")
# gomp_fit <- fit_bayesian_growth(..., model = "g")
# logis_fit <- fit_bayesian_growth(..., model = "l")

# CW mortality from each growth model
mort_from_vb <- get_stochastic_mortality(
  method = "CW", growth_fit = vb_fit, sex = 1, print_plot = FALSE
)
mort_from_gomp <- get_stochastic_mortality(
  method = "CW", growth_fit = gomp_fit, sex = 1, print_plot = FALSE
)
mort_from_logis <- get_stochastic_mortality(
  method = "CW", growth_fit = logis_fit, sex = 1, print_plot = FALSE
)

combined_growth <- rbind(
  mort_from_vb$Summary[, growth_model := "von Bertalanffy"],
  mort_from_gomp$Summary[, growth_model := "Gompertz"],
  mort_from_logis$Summary[, growth_model := "Logistic"]
)

ggplot(combined_growth, aes(x = age_round, color = growth_model, fill = growth_model)) +
  geom_ribbon(aes(ymin = M_lower, ymax = M_upper), alpha = 0.2, color = NA) +
  geom_line(aes(y = M_median), linewidth = 1) +
  labs(x = "Age (years)", y = "Natural Mortality (M)",
       title = "CW Mortality by Growth Model",
       subtitle = "All derived from same biological milestones") +
  theme_bw() +
  theme(legend.position = "top")
```

If the three curves are similar, mortality estimates are robust to growth model choice. Divergence indicates genuine model uncertainty that should be reported.

## Modular Mortality Functions

For applications outside the Monte Carlo framework --- quick calculations, integration into custom analyses, or pedagogical exploration --- vitalBayes exports the individual mortality model functions:

```{r modular}
ages <- seq(0.5, 30, by = 0.5)

# Chen-Watanabe (single-phase and two-phase)
M_cw_single <- M_chen_watanabe_L0(
  age = ages, Linf = 100, L0 = 25, k = 0.08,
  two_phase = FALSE
)

M_cw_senesc <- M_chen_watanabe_L0(
  age = ages, Linf = 100, L0 = 25, k = 0.08,
  two_phase = TRUE, tmat = 12, late_model = "gompertz"
)

# Peterson-Wroblewski
lw_func <- function(L) 0.0001 * L^3.1
M_pw <- M_peterson_wroblewski(
  age = ages, Linf = 100, L0 = 25, k = 0.08,
  lw_fun = lw_func, growth_model = "vb"
)

# Lorenzen growth-based
M_lor <- M_lorenzen(
  age = ages, Linf = 100, L0 = 25, k = 0.08,
  weight_based = FALSE
)

# Scale any of them
M_scaled <- scale_mortality(M_cw_senesc, M_target = NULL, tmax = 35, p = 0.001)
```

## Downstream Integration: Survival Simulation

Mortality schedules from `get_stochastic_mortality()` feed directly into `simulate_survivorship()` for cohort survival analysis:

```{r downstream}
surv <- simulate_survivorship(
  mc_object = mort,
  n         = 50000,    # Starting cohort size
  n_iter    = 2000,     # Simulation iterations
  mode      = "random"  # Sample parameter sets randomly
)

# Key survival metrics
surv$Aggregate$Age_of_Death
#>     mean    sd  lower  upper
#>    <num> <num>  <num>  <num>
#> 1:  8.42  2.31   4.89  13.21

surv$Aggregate$Survival_to_tmat
#>     mean      sd   lower   upper
#>    <num>   <num>   <num>   <num>
#> 1: 0.312  0.0891   0.168   0.512
```

See `vignette("survivorship_simulation")` for comprehensive coverage of survival analysis options, visualization, and interpretation.

## Troubleshooting

| Issue | Possible Cause | Solution |
|-------|----------------|----------|
| Negative mortality values | CW two-phase Taylor expansion breakdown | Function automatically caps at 0; consider single-phase or different `late_model` |
| Very wide uncertainty bands | Weak growth model posterior | Check growth model convergence and parameter correlations |
| Unrealistic tmax estimates | `Linf_factor` too extreme | Reduce from 0.99 to 0.95 for exploratory analysis |
| Missing tmat error | CW needs age-at-maturity | Provide `maturity_fit` or manual `tmat` parameter |
| `lw_fun` error | PW/Lorenzen weight-based needs length-weight | Provide function: `lw_fun = function(L) a * L^b` |
| CW and PW disagree strongly | Different theoretical assumptions | Report both; consider which assumptions best match your species |

## Reporting Results

When reporting mortality estimates in publications, include:

1. **Model specification**: Which mortality model(s), growth model used, scaling approach, and parameter sources.
2. **Growth parameter source**: vitalBayes fit (specify model: VB, Gompertz, or Logistic) or literature values.
3. **Key metrics**: Mean $M$, age-specific $M$ at key life stages (e.g., neonate, sub-adult, adult), $t_{max}$ estimate.
4. **Uncertainty**: 95% credible intervals throughout, noting whether these reflect joint posterior sampling or independent parameter variation.

Example methods text:

> "Natural mortality was estimated using the Chen-Watanabe model (Chen & Watanabe, 1989) with Gompertz senescence, implemented via the vitalBayes package (v1.0.0; Author, Year). Growth parameters were drawn from the joint posterior of a Gompertz growth model (selected by LOO-CV), with VB-equivalent $k$ computed from biological milestones following the approach described in [vignette reference]. Mortality schedules were scaled to 0.1% survival at estimated $t_{max}$ and propagated through 2,000 Monte Carlo simulations. Median age-specific mortality with 95% credible intervals is reported."

## References

Chen, S., & Watanabe, S. (1989). Age dependence of natural mortality coefficient in fish population dynamics. *Nippon Suisan Gakkaishi*, 55(2), 205--208.

Lorenzen, K. (1996). The relationship between body weight and natural mortality in juvenile and adult fish. *Journal of Fish Biology*, 49(4), 627--642.

Lorenzen, K. (2022). Size- and age-dependent natural mortality in fish populations. *Fisheries Research*, 255, 106454.

Peterson, I., & Wroblewski, J. S. (1984). Mortality rate of fishes in the pelagic ecosystem. *Canadian Journal of Fisheries and Aquatic Sciences*, 41(7), 1117--1120.

Then, A. Y., Hoenig, J. M., Hall, N. G., & Hewitt, D. A. (2015). Evaluating the predictive performance of empirical estimators of natural mortality rate using information on over 200 fish species. *ICES Journal of Marine Science*, 72(1), 82--92.

## See Also

- `vignette("chen_watanabe_reparameterization")` --- Mathematical derivation of the CW reparameterization and $G(t)$ framework
- `vignette("survivorship_simulation")` --- Cohort survival analysis using mortality schedules
- `vignette("fit_bayesian_growth")` --- Growth model fitting that feeds the mortality workflow
- `vignette("fit_bayesian_maturity")` --- Maturity estimation for $t_{mat}$ and $L_{mat}$
